---
title: "Accelerating co-ordinate ascent updates for linear regression using DAAREM"
date: June 4, 2019
site: workflowr::wflow_site
output: workflowr::wflow_html
---

A small script to illustrate application of the DAAREM method for
computing maximum-likelihood estimates of mixture proportions in a
mixture model.

```{r knitr, echo=FALSE}
knitr::opts_chunk$set(comment = "#",results = "hold",collapse = TRUE,
                      fig.align = "center")
```

## Set up environment

Load some packages and function definitions used in the example below.

```{r load-pkgs, warning=FALSE, message=FALSE}
library(ggplot2)
library(cowplot)
library(daarem)
source("../code/misc.R")
source("../code/mixem.R")
```

## Load data set

TO DO: Add text here.

```{r load-data}
load("../data/mixdata.RData")
n <- nrow(L)
m <- ncol(L)
cat(sprintf("Loaded %d x %d data matrix.\n",n,m))
```

Set the initial estimate of the mixture proportions.

```{r}
x0 <- rep(1/m,m)
```

## Run basic EM updates

Compute maximum-likelihood estimates of the mixture proportions by
running 200 iterations of the EM updates.

```{r fit-em}
cat("Fitting mixture model with basic EM method.\n")
out <- system.time(fit1 <- mixem(L,x0,numiter = 200))
f1  <- mixobjective(L,fit1$x)
cat(sprintf("Computation took %0.2f seconds.\n",out["elapsed"]))
cat(sprintf("Log-likelihood at EM estimate is %0.12f.\n",f1))
```

## Run accelerated EM

Re-run the EM updates, this time using DAAREM to accelerate
convergence toward the solution.

```{r fit-daarem}
out <- system.time(fit2 <- mixdaarem(L,x0,numiter = 200))
f2  <- mixobjective(L,fit2$x)
cat(sprintf("Computation took %0.2f seconds.\n",out["elapsed"]))
cat(sprintf("Objective value at DAAREM estimate is %0.12f.\n",f2))
```

## Plot improvement in solution over time

TO DO: Add text here.

```{r plot-iter-vs-objective, fig.height=4, fig.width=6}
f    <- mixobjective(L,x)
pdat <-
  rbind(data.frame(iter = 1:200,dist = f - fit1$value,method = "EM"),
        data.frame(iter = 1:200,dist = f - fit2$value,method = "DAAREM"))
p <- ggplot(pdat,aes(x = iter,y = dist,col = method)) +
  geom_line(size = 1) +
  scale_y_continuous(trans = "log10",breaks = 10^seq(-4,4)) +
  scale_color_manual(values = c("darkorange","dodgerblue")) +
  labs(x = "iteration",y = "distance from solution")
print(p)
```
